Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2023-01-03 revision/review_response.qmd

* __Response__: Thank you for your detailed comments on our manuscript.  We have made every effort to address your concerns noted below and are confident this has improved the draft.

Abstract:

While the paper is focused on the method and spatial classification to identify areas for potential restoration, this is not represented in the Abstract. Instead, the Abstract indicates a focus on the target-setting approach and emphasizes how sea-level rise, climate change, and watershed development are incorporated into this approach. The Abstract should rather focus on the main results of the paper – which was the identification of the different areas and changes over time as measured by applying the method. That is not to say that the entire Abstract is unsuitable, rather that it needs to be amended to better reflect the contents of the paper, rather than a summary of the approach.

* __Response__: Thanks for this suggestion.  We have revised the abstract to include more relevant content from the paper. The entire abstract from line 17 down was edited:

     "Although significant gains in subtidal habitats have been observed, expansion of mangroves into salt marshes and loss of native upland habitats to development highlights the need to target these locations for restoration. The long-term loss of potentially restorable lands to both coastal and upland development further underscores the diminishing restoration opportunities in the watershed. The established targets and goals identified habitats to maintain at their present level (e.g., mangroves) and those that require additional progress (e.g., oyster bars) based on past trends and an expected level of effort given the restoration history of the region. The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas. Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the targets and goals, with methods for repeatable analyses also available using an open source workflow."

@@

L127: At first mention of the 5-foot contour please provide an estimate in metres for comparison. This will also make for a more straightforward understanding of the SLR projection, which is provided in cm (L132).

* __Response__: This sentence was revised as follows "...location relative to the 5-foot contour (~1.5m elevation) that covers an area of land...".

L130-132: The SLR trend and projections to 2100 are stated here, but it is not clear whether these have been incorporated into the 2030 / 2050 targets / goals for restoration. In the Introduction (L83-86), it is stated that the approach will consider future stressors, including SLR. A spatial representation of how habitat / land cover may change by 2030 / 2050 under SLR is however not part of the Results. Were the SLR projections incorporated spatially? And did this impact the potential restoration areas? For example, are transitions from mangrove to tidal flat / seagrass included in the targets / goals when considering SLR? Please provide some additional information on whether this was incorporated and how it was done. This is needed to justify the statement in L83-86.

* __Response__: The expected effects of sea level rise are implicitly accounted for in the 5-foot contour that was used to differentiate the opportunity areas in the coastal margin from those in the upland.  By placing emphasis on these areas, it is anticipated that restoration practitioners can target these locations differently based on the diminishing opportunities in these areas.  For example, figures 5 and 6 distinguish these areas from the other coastal strata.  This area also accounts for the projected increases in sea level rise in @Burke19.  The following text was added to line 132 to make this clear:

@@

     "The coastal stratum within the 5-foot contour is used to better identify and prioritize coastal habitats at risk of landward migration and coastal development given that it includes the area of land within the sea level rise projections."

L154-165: Please indicate the type and scale of the spatial datasets for the habitat. Are these raster datasets? What is the grid resolution? Are they obtained from Landsat, or other remotely-sensed products? Providing detail on the input datasets will greatly assist any effort to replicate this method in another region.

* __Response__: Both the subtidal and inter/supratidal data layers are vector products produced by photo-interpretation.  Details were added as follows.

     For the subtidal layers, details were added starting on line 150: "These data include vector polygon coverages of the major subtidal habitats in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.  Accuracy assessments of the photo-interpreted map included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.  The minimum mapping unit for seagrass polygons is reported as 0.25 acres."

@@

     For the inter/supratidal layers, details were added on line 158: "Similar methods as the subtidal habitats described above are used for the intertidal and supratidal coverage maps, although at a slightly higher spatial resolution (1:12,000)."

L170-175: Please provide more detail on how the habitat coverage change analysis was performed. Was this a simple calculation of total area change per habitat type derived from pixel count / polygon area? Or, were the maps for different years overlayed and a spatial algorithm applied to calculate differences? Please specify how the change from one habitat to another was estimated in order to generate the alluvial diagrams.

* __Response__: Additional detail was provided as follows: "Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.  For example, the area that remained as native uplands between the 1990 and 2017 intertidal and supratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.  This process was repeated for all native habitats, including developed and restorable lands (described below)."

@@

L177-188: While I understand the justification to differentiate between “restoration” and “enhancement” activities, I would suggest the authors re-consider the terminology and definitions they have provided. The term “restoration” is well-known, with the definition provided by the Society of Ecological Restoration becoming mainstreamed, particularly under the “Decade for Ecosystem Restoration”. The SER definition for restoration from the international principles and standards (https://www.ser.org/page/SERStandards): “the process of assisting the recovery of an ecosystem that has been degraded, damaged or destroyed. (Ecosystem restoration is sometimes used interchangeably with ecological restoration, but ecological restoration always addresses biodiversity conservation and ecological integrity, whereas some approaches to ecosystem restoration may focus solely on the delivery of ecosystem services.)” Therefore, it is advised that the authors do not provide their own definition for restoration to only refer to activities where earthwork is involved to reshape the land. Rather, it would be better to state what type of restoration activity has been conducted, instead of using the holistic / encompassing term “restoration” to only refer to a specific subset of activities. Given the definition provided by the authors, it needs to be clarified if all areas indicated as “potential restorable’ would therefore require earthwork in order to revert to natural habitat? Or in this case does “restorable” encompass the greater suite of possible activities?

* __Response__: Thank you for the clarification and we agree that adopting a standardized and more widely accepted definition is more appropriate. However, the projects in the database have already been classified as restoration/enhancement based on the previous definition and this distinction is useful to local restoration practitioners that are familiar with the projects.  For clarity, we have included the SER definition as a general description and amended the text to better justify the previous description.  The addition starting on line 179 is as follows:

@@

     "Here and throughout, restoration describes the process of assisting the recovery of an ecosystem that has been degraded, damaged, or destroyed [@Gann19].  Restoration projects in the database were also those that involved earthwork to reshape the land or the addition of structural elements (e.g., rock). This distinct categorization is useful for restoration practitioners familiar with the projects in the region."

L206-207 and Table1: Is there a difference in using the term “reservation” when referring to the coastal areas, vs. the term “conservation” being used for the upland areas? Looking at Table 4, there are coastal habitats in conservation areas. Is this to just designate specifically areas that are within the 5ft contour, regardless of protection status – this was my understanding from Table 1, but it would be beneficial to have it stated more clearly in the text.

* __Response__: Apologies for the confusion.  The reservation areas are those in the coastal stratum that are currently not in existing or proposed conservation areas.  The sentence on line 211 was revised as follows:

@@

     "The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum and do not occur in existing or proposed conservation areas (described in the following paragraph)."

L217-219: Again, please provide some detail on the spatial resolution of these datasets. Were there any issues / challenges with alignment with spatial datasets from other sources? How were these overcome?

* __Response__: Details on these data were added as follows: "The FNAI data are Florida Managed Areas as vector polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.  The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000."

@@

L263-266: Could changes in seagrass also be attributed to improved ground-truthing and photointerpretation? Perhaps some additional information on the input datasets, such as how the areas were estimated, could provide some insight (as suggested to present in a table above in Methods section).

* __Response__: The change in seagrasses in Tampa Bay over the last three decades is well-documented [e.g., @Greening14; @Sherwood17].  The documentation of this recovery has been made possible primarily from efforts of the Southwest Florida Water Management District in providing accurate and comparable coverage maps on an approximate biennial basis since the late 1980s (as described in the data sources).  Although some minor changes in the FLUCCS codes have occurred during this period, the methods for documenting seagrasses have been consistent. We are confident that these changes are not a product of changes in methodologies.  We hope that our addition to the text describing more detail on the data sources and methods has provided some clarity.

L285-295: This is valuable information, and impressive that the authors were able to collate and document the restoration projects that have been carried out across different agencies and partnerships. This information will be very important for monitoring and reporting on progress for targets going forward, so it is a critical step to have synthesized the work that has already been carried out. An additional useful output would be a spatial representation of these projects so that the map could be updated as more projects are completed in the future (but this would be beyond the scope of this paper).

* __Response__: Yes, we agree that the effort to create and synthesize these data was one of the more useful products resulting from this work. Current discussions have focused on how to maintain the integrity of these data as additional work is completed in the future. The limitations discussed in the second paragraph of section 4.3 alludes to these challenges. A link to the online location of this database, that includes a map, was added to line 188 in the methods (<https://www.tampabay.wateratlas.usf.edu/restoration/>).

@@

L309-310: Should this read “(coastal reservation restorable)”?

* __Response__: Corrected.

L324-335: In this section describing the targets and goals, I anticipated more detail on the potential climate change effects on the habitat map. As mentioned above in a previous comment, if this has been included then please provide a clearer description of the approach. But for example, the distribution of the intertidal habitats could be different by 2050 under SLR, and therefore adjacent upland area may be lost / transitioned into intertidal habitat. The authors have mentioned landward migration and the possibility of coastal squeeze due to infrastructure in the Discussion, but I expected it to be more clearly incorporated into the Results, given the framing of the paper in the Abstract and Introduction. If this is not the case then those previous sections need to be revisited.

* __Response__: We understand the general concern that we have not included explicit projections of how land use and cover may change with climate change and development.  The methods we have used implicitly account for these factors and are our best estimate at what can be reasonably achieved by 2030 and 2050 based on the assessment.  We hope that revisions to some of the earlier comments have also made this more clear (e.g., lines 130 - 132).  We have added the following text to this section for clarity:

@@

The authors have provided a suitable Discussion to contextualise the results of the paper.

Tables:

Table3: Suggest that the heading in the second column be changed to “Landcover Category” or something similar, instead of “Habitat Type” as this column includes “Developed” and “Restorable”, which would not typically be considered as habitats.

* __Response__: The column header was changed to "Land cover Category".  The table caption was also changed.

@@

Table4: There are some discrepancies in the table, or perhaps a sentence is needed in the title to explain why the current extent is not equal to existing conservation + proposed conservation lands. For example, Mangroves current extent = 6276, existing conservation + proposed conservation = (4516+1604) = 6120. Does this indicate an additional 156 ha of mangrove that is not within a conservation area? Also, check the rounding up of the area estimates as there is a discrepancy in values presented for total restoration opportunity = existing conservation lands restoration opportunity + proposed conservation lands restoration opportunity.

* __Response__: Correct, the current extent is the sum of existing and proposed conservation, plus those not in conservation.  A sentence was added to the caption. The rounding issues for total restoration opportunity were also fixed.

Table5: This table is arguably the most important result from the paper. The only query I have is whether the authors have considered how the map would look if these targets / goals were to be achieved. For example, the target for total intertidal to increase by 61 hectares in a mosaic – how does this factor in SLR?

* __Response__: Please see our response to your previous comment on lines 324-335. One could potentially understand how the land use and cover may change if the targets and goals were met by viewing the restoration potential in Figure 6.  However, the restoration potential is general and does not provide specific recommendations on which native habitats to restore, i.e., tidal wetland is indicated and not mangrove forests, salt barrens, or salt marshes.  Additional data would be needed to determine the appropriate type of tidal wetland and we leave that to the restoration practitioners to conduct site-level assessments for planned restorations.  Also, our responses to the other reviewer addresses this concern regarding the true restoration potential of wetlands.

@@

It is also not immediately clear from the table how the values for the 2030 targets and 2050 goals are calculated. For example, for coastal uplands if 513 ha is the total restoration opportunity, then how was 1689 ha estimated for the 2050 target (if the current extent is 1446 ha)?

* __Response__: The targets and goals are explained by the rationale in the final column.  These values are not inclusive of the total restoration opportunity, rather they are set by the expectation of what can be restored during these time periods based on past restoration effort and trends.  For example, the coastal uplands target for 2030 is the current extent plus 61 hectares (1446 ha + 61 ha = 1507 ha) and the goal for 2050 is 1507 ha plus 61 ha increased by 20 each out decade (1507 ha + 81 ha + 101 ha = 1689 ha).  The paragraph on line 324 was edited to make this more clear.

### 2023-01-03 manu-draft.qmd
finished first round of revisions and proofed

Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise. To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of native habitats, and anticipated future impacts from coastal stressors. A restoration framework for native habitats was created for the Tampa Bay watershed that identifies current opportunities and establishes short-term (2030) targets and long-term (2050) goals. The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region. Although significant gains in subtidal habitats have been observed, expansion of mangroves into salt marshes and loss of native upland habitats to development highlights the need to target these locations for restoration.  The long-term loss of potentially restorable lands to both coastal and upland development further underscores the diminishing restoration opportunities in the watershed.  The established targets and goals identified habitats to maintain at their present level (e.g., mangroves) and those that require additional progress (e.g., oyster bars) based on past trends and an expected level of effort given the restoration history of the region.  The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas. The restoration potential for coastal habitats on existing conservation lands was `r areas$potential[['Tidal Wetlands']]$areaha` for tidal wetlands and `r areas$potential[['Coastal Uplands']]$areaha` for coastal uplands.  Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals. Methods for repeatable analyses are also available using an open source workflow.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].  These data include vector polygons coverages of the major subtidal habitat types in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.  Accuracy assessments of the photo-interpreted map product included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.  The minimum mapping unit for seagrass polygons is reported as 0.25 acres. Historical datasets for subtidal habitats using identical methods began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

@@

Finally, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types. This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats). Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.  For example, the area that remained as native uplands between the 1990 and 2017 intertidal and suptratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.  This process was repeated for all native habitats, including developed and restorable lands (described below).  The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].

@@

The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition. To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory (FNAI) and permit databases of conservation and drainage easements [@FNAI20]. The FNAI data are Florida Managed Areas as vector-based polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.  The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper source maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000. Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable. This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow.

@@

Table 5 shows the targets and goals identified through this analysis and the associated rationale. For example, the targets and goals are established based on the current extent and the restoration potential. If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity. Conversely, other habitats with no identified restoration opportunity, or with sufficient current extents (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and further restoration will only increase resilience. Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.

@@

```{r}
#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal landcover in Tampa Bay.  Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017."
acretab
```

>>>>

Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise. To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of native habitats, and anticipated future impacts from coastal stressors. A restoration framework for native habitats was created for the Tampa Bay watershed that identifies current opportunities and establishes short-term (2030) targets and long-term (2050) goals. The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region. Although significant gains in subtidal habitats have been observed, expansion of mangroves into salt marshes and loss of native upland habitats to development highlights the need to target these locations for restoration.  The long-term loss of potentially restorable lands to both coastal and upland development further underscores the diminishing restoration opportunities in the watershed.  The established targets and goals identified habitats to maintain at their present level (e.g., mangroves) and those that require additional progress (e.g., oyster bars) based on past trends and an expected level of effort given the restoration history of the region.  The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas. Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the targets and goals, with methods for repeatable analyses also available using an open source workflow.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].  These data include vector polygon coverages of the major subtidal habitats in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.  Accuracy assessments of the photo-interpreted map included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.  The minimum mapping unit for seagrass polygons is reported as 0.25 acres. Historical datasets for subtidal habitats using identical methods began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

@@

Finally, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types. This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats). Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.  For example, the area that remained as native uplands between the 1990 and 2017 intertidal and supratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.  This process was repeated for all native habitats, including developed and restorable lands (described below).  The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].

@@

The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition. To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory (FNAI) and permit databases of conservation and drainage easements [@FNAI20]. The FNAI data are Florida Managed Areas as vector polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.  The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000. Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable. This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow.

@@

Table 5 shows the targets and goals identified through this analysis and the associated rationale. For example, the targets and goals are established based on the current extent and informed by the restoration potential. If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent relative to the restoration opportunity, past trends, and anticipated effort. Conversely, other habitats with no identified restoration opportunity, or with sufficient current extents (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and further restoration will only increase resilience. The proposed targets and goals do not represent the current extent plus restoration potential for these reasons. Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.

@@

```{r}
#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal land cover in Tampa Bay.  Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017."
acretab
```
